---
title: "*ctsem* Advice Updated"
author: "Blain Morin"
date: "10/11/2021"
output: pdf_document
header-includes:
   - \usepackage{geometry}
   - \usepackage{placeins}
   - \usepackage{amsmath} 
   - \usepackage{bm}
   - \usepackage{lscape}
   - \usepackage{rotating}
   - \newcommand{\vect}[1]{\boldsymbol{\mathbf{#1}}}
   - \newcommand{\blandscape}{\begin{landscape}}
   - \newcommand{\elandscape}{\end{landscape}}
---

# One-sentence summary

We are interested in measuring the latent capacity and automony of federal agencies over time.

# Why capacity and autonomy?

* *Capacity*: Ability of the state to induce residents, firms, and organizations to act in ways they would otherwise not.

* *Autonomy*: Measures the politicization of the bureaucracy. How much control do elected officials have over bureaucrats?

* These characteristics may be associated with governance outcomes.

  * Ex.) Are capacity and automony of the Environmental Protection Agency related to their enforment work?

* Is there hetergeneity across agencies?

  * Ex.) Has the environmental state been hollowed out during the last few years?

* Call in the political science literature for better conceptual clarity and measurement of these latent characteristics.

* Prior work is cross-sectional. Need to understand how capacity and autonomy interact. 

# Data

* Have employment data for all federal agencies for about the past 40 years

* Can observe characteristics such as years of service, education level, and salary of individual employees at every agency

* Group these observations to the agency,year to get measures such as \% of employees with a master's degree or higher or average length of service (*our manifest variables*).

* Able to replicate others' work on measurement, or propose our own list of relevant manifest variables. 


# Simple Test Set

* Starting simple, I first try to model capacity using two manifest variables, *\% of employees at an agency who have a master's degree or higher* and *median salary at the agency*   
* Only larger Environmental Agencies where the number of employees > 5000
* 30 years
* See Table 1: we have 11 agencies, each with a fully balanced 30 observations

\FloatBarrier

```{r setup, include=FALSE}

### Document Settings
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, cache = TRUE)

### Packages
library(tidyverse)
library(ctsem)
library(rstan)
library(kableExtra)

### Raw Data
fed = read.csv("https://www.dropbox.com/s/xtt4emmz6txtbun/fed_agency_capacity_autonomy.csv?dl=1")

### Set Seed
set.seed(3710)


```

```{r}

# Choose Year Range
startyear = 1981 # Needs to be >=1974
endyear = 2010 # Needs to be <=2019

# Choose Agency Type
agencytype = c("Natural Resources and Environment")
# agencytype = c("Natural Resources and Environment",
#                "Health")

# Choose Manifest Variables 
regressors = c("ma_pct", "med_sal_")
# names(fed)

# Choose Minimum Employee Size
minemployee = 5000


```

```{r}

# Data Clean
df = fed %>%
  filter(yr >= 1974) %>%
  filter(med_sal_ > 0) %>% 
  filter(n > minemployee) %>%
  drop_na(med_sal_) %>% ### Am dropping NA here, but may not need to with ctsem
  filter(AGYSUB != "TOTL") %>% ### Removes Total Rows 
  filter(!grepl("NET", agy_full)) ### Removes any total agency counts, ie only individual agencies are left

# Data Process
years = startyear:endyear
dff = df %>%
  filter(agy_typ %in% agencytype) %>%
  filter(yr %in% startyear:endyear) %>%
  select(regressors, yr, AGYSUB, agy_full) %>%
  mutate_at(regressors, scale) %>%
  drop_na() ### Again, dropping NA here, but may not need to

check = dff %>%
  group_by(agy_full, AGYSUB) %>%
  summarise(n_obs = n())

kbl(check, caption = "Agencies and Number of Observations for 1991 - 2010", booktabs = TRUE, linesep = "")

```

# Discussion: Identifiablility
Prior experience is nudging me towards the idea in order to make the latent solution identifiable, that we need to set the *scale* and *location* of the latent variable. In order to set the scale, I force the first $\Lambda_{1,1} = 1$. In order to set the location, I force the first $manifestmean[1,1] = 0$.

* Question: Is this identification correct?
* Since the manifest variables are scaled, should I force all manifest means in the estimation to 0?

# Discussion: Order of Regressors

One possible problem is that the order of the regressors seems to impact the model results. For example, in the following section, I run the same model but only change the order of the regressors:

## Model 1: Median Salary First, MA \% second 

```{r}

## Make Strings for the model
## based on the number of regressors
lambdas = c(1)
manifest = c(0)
for (i in 2:length(regressors)) {
  
  temp = paste0("lambda", i)
  temp2 = paste0("manifestmean", i)
  lambdas[i] = temp
  manifest[i] = temp2
}

```



```{r, echo = TRUE}

regressors = c("med_sal_", "ma_pct")

# The is the model specification
model = ctModel(type='stanct', 
               
               LAMBDA=matrix(lambdas,
                             nrow = length(regressors), 
                             ncol = 1), # Forcing the first factor load to be 1
               
               n.manifest=length(regressors),
               
               manifestNames = regressors,
               
               MANIFESTMEANS = matrix(manifest,
                                      nrow = length(regressors), 
                                      ncol = 1), # Forcing the first mean to be 0
               
               n.latent=1,
               
               latentNames=c('capacity'),
               
               CINT = matrix('cint'),
               
               id = "agy_full",
               
               time = "yr")

```

\FloatBarrier

```{r}

fit = ctStanFit(datalong = dff,
                ctstanmodel = model,
                optimize = TRUE,
                nopriors = FALSE)

```

\FloatBarrier

```{r, results='asis'}

cat(paste0('\\begin{sidewaysfigure}'))
cat(ctModelLatex(fit,textsize = 'small',compile = FALSE, equationonly = TRUE))
cat('\\caption{Both on}
\\label{fig:ctmodellatexout}
\\end{sidewaysfigure}') 

```