---
title: "*ctsem* Advice Updated"
author: "Blain Morin"
date: "10/11/2021"
output: pdf_document
header-includes:
   - \usepackage{geometry}
   - \usepackage{placeins}
   - \usepackage{amsmath} 
   - \usepackage{bm}
   - \usepackage{lscape}
   - \usepackage{rotating}
   - \newcommand{\vect}[1]{\boldsymbol{\mathbf{#1}}}
   - \newcommand{\blandscape}{\begin{landscape}}
   - \newcommand{\elandscape}{\end{landscape}}
---

# One-sentence summary

We are interested in measuring the latent capacity and automony of federal agencies over time.

# Why capacity and autonomy?

* *Capacity*: Ability of the state to induce residents, firms, and organizations to act in ways they would otherwise not.

* *Autonomy*: Measures the politicization of the bureaucracy. How much control do elected officials have over bureaucrats?

* These characteristics may be associated with governance outcomes.

  * Ex.) Are capacity and automony of the Environmental Protection Agency related to their enforment work?

* Is there hetergeneity across agencies?

  * Ex.) Has the environmental state been hollowed out during the last few years?

* Call in the political science literature for better conceptual clarity and measurement of these latent characteristics.

* Prior work is cross-sectional. Need to understand how capacity and autonomy interact. 

# Data

* Have employment data for all federal agencies for about the past 40 years

* Can observe characteristics such as years of service, education level, and salary of individual employees at every agency

* Group these observations to the agency,year to get measures such as \% of employees with a master's degree or higher or average length of service (*our manifest variables*).

* Able to replicate others' work on measurement, or propose our own list of relevant manifest variables. 


# Simple Test Set


* Starting simple, I first try to model capacity using two manifest variables, *\% of employees at an agency who have a master's degree or higher* and *median salary at the agency*  

* Only larger Environmental Agencies where the number of employees > 5000

* 30 years

* See Table 1: we have 11 agencies, each with a fully balanced 30 observations

\FloatBarrier

```{r setup, include=FALSE}

### Document Settings
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, cache = TRUE)

### Packages
library(tidyverse)
library(ctsem)
library(rstan)
library(kableExtra)

### Raw Data
fed = read.csv("https://www.dropbox.com/s/xtt4emmz6txtbun/fed_agency_capacity_autonomy.csv?dl=1")

### Set Seed
set.seed(3710)


```

```{r}

# Choose Year Range
startyear = 1981 # Needs to be >=1974
endyear = 2010 # Needs to be <=2019

# Choose Agency Type
agencytype = c("Natural Resources and Environment")
# agencytype = c("Natural Resources and Environment",
#                "Health")

# Choose Manifest Variables 
regressors = c("ma_pct", "med_sal_")
# names(fed)

# Choose Minimum Employee Size
minemployee = 5000


```

```{r}

# Data Clean
df = fed %>%
  filter(yr >= 1974) %>%
  filter(med_sal_ > 0) %>% 
  filter(n > minemployee) %>%
  drop_na(med_sal_) %>% ### Am dropping NA here, but may not need to with ctsem
  filter(AGYSUB != "TOTL") %>% ### Removes Total Rows 
  filter(!grepl("NET", agy_full)) ### Removes any total agency counts, ie only individual agencies are left

# Data Process
years = startyear:endyear
dff = df %>%
  filter(agy_typ %in% agencytype) %>%
  filter(yr %in% startyear:endyear) %>%
  select(regressors, yr, AGYSUB, agy_full) %>%
  mutate_at(regressors, scale) %>%
  drop_na() ### Again, dropping NA here, but may not need to

check = dff %>%
  group_by(agy_full, AGYSUB) %>%
  summarise(n_obs = n())

kbl(check, caption = "Agencies and Number of Observations for 1991 - 2010", booktabs = TRUE, linesep = "")

```

# Discussion: Identifiablility
Prior experience is nudging me towards the idea in order to make the latent solution identifiable, that we need to set the *scale* and *location* of the latent variable. In order to set the scale, I force the first $\Lambda_{1,1} = 1$. In order to set the location, I force the first $manifestmean[1,1] = 0$.

* Question: Is this identification correct?
* Since the manifest variables are scaled, should I force all manifest means in the estimation to 0?

# Discussion: Order of MAnifest Variables

One possible problem is that the order of the manifest seems to impact the model results unecpectedly. For example, in the following section, I run the same model but only change the order of the manifest variables:

## Model 1: Median Salary First, MA \% second 

```{r}

## Make Strings for the model
## based on the number of regressors
lambdas = c(1)
manifest = c(0)
for (i in 2:length(regressors)) {
  
  temp = paste0("lambda", i)
  temp2 = paste0("manifestmean", i)
  lambdas[i] = temp
  manifest[i] = temp2
}

```



```{r, echo = TRUE}

regressors = c("med_sal_", "ma_pct")

# The is the model specification
model = ctModel(type='stanct', 
               
               LAMBDA=matrix(lambdas,
                             nrow = length(regressors), 
                             ncol = 1), # Forcing the first factor load to be 1
               
               n.manifest=length(regressors),
               
               manifestNames = regressors,
               
               MANIFESTMEANS = matrix(manifest,
                                      nrow = length(regressors), 
                                      ncol = 1), # Forcing the first mean to be 0
               
               n.latent=1,
               
               latentNames=c('capacity'),
               
               CINT = matrix('cint'),
               
               id = "agy_full",
               
               time = "yr")

```

\newpage

\FloatBarrier

```{r}

fit = ctStanFit(datalong = dff,
                ctstanmodel = model,
                optimize = TRUE,
                nopriors = FALSE)

```

\FloatBarrier

```{r, results='asis'}

cat(paste0('\\begin{sidewaysfigure}'))
cat(ctModelLatex(fit,textsize = 'small',compile = FALSE, equationonly = TRUE))
cat('\\caption{Salary First, Education Second}
\\label{fig:ctmodellatexout}
\\end{sidewaysfigure}') 

```

\FloatBarrier

### Model 1 Kalmman Plot, median salary first

\FloatBarrier

```{r, echo = TRUE, fig.height=5, fig.width=8}

ctKalman(fit, 
         plot = TRUE,
         subjects = (1:length(unique(dff$AGYSUB))), # Plots all agencies
         kalmanvec = c("etasmooth", "y", "ysmooth"),
         timestep = .5) 

```

As we see in the Kalman plot, capacity seems to follow the med_sal_ trend. Most of the error bands are in ma_pct. This pattern changes, however, if I change the order of the manifest variables.

\FloatBarrier

## Model 2: MA\% First, Median Salary Second

\FloatBarrier

```{r}

## Make Strings for the model
## based on the number of regressors
lambdas = c(1)
manifest = c(0)
for (i in 2:length(regressors)) {
  
  temp = paste0("lambda", i)
  temp2 = paste0("manifestmean", i)
  lambdas[i] = temp
  manifest[i] = temp2
}

```


```{r, echo = TRUE}

# Switch order of regressors
regressors = c("ma_pct", "med_sal_")

# The is the model specification
model = ctModel(type='stanct', 
               
               LAMBDA=matrix(lambdas,
                             nrow = length(regressors), 
                             ncol = 1), # Forcing the first factor load to be 1
               
               n.manifest=length(regressors),
               
               manifestNames = regressors,
               
               MANIFESTMEANS = matrix(manifest,
                                      nrow = length(regressors), 
                                      ncol = 1), # Forcing the first mean to be 0
               
               n.latent=1,
               
               latentNames=c('capacity'),
               
               CINT = matrix('cint'),
               
               id = "agy_full",
               
               time = "yr")

```

```{r}

fit2 = ctStanFit(datalong = dff,
                ctstanmodel = model,
                optimize = TRUE,
                nopriors = FALSE)

```

\newpage

\FloatBarrier

```{r, results='asis'}

cat(paste0('\\begin{sidewaysfigure}'))
cat(ctModelLatex(fit2,textsize = 'small',compile = FALSE, equationonly = TRUE))
cat('\\caption{Education First, Salary Second}
\\label{fig:ctmodellatexout1}
\\end{sidewaysfigure}') 

```


\FloatBarrier

\newpage

### Model 2: Kalman Plot, MA \% first

\FloatBarrier

```{r, echo = TRUE, fig.height=5, fig.width=8}

ctKalman(fit2, 
         plot = TRUE,
         subjects = (1:length(unique(dff$AGYSUB))), # Plots all agencies
         kalmanvec = c("etasmooth", "y", "ysmooth"),
         timestep = .5) 

```

Counter to model 1, capacity now follows the MA_pct trend instead of the med_sal_ trend. The uncertainty bars have moved from education to salary as well. My thought is that the order of the manifest variables shouldn't impact the capacity trend, or affect the se on the manifest plots. 

# Add autonomy (second latent variable)

Eventually, would like to build up to adding a second latent variable, autonomy. We are also interested in the cross-effect. How does the autonomy at time $t$ affect the capacity of an agency at future time points?

```{r}

# Choose Year Range
startyear = 1991 # Needs to be >=1974
endyear = 2010 # Needs to be <=2019

# Choose Agency Type
agencytype = c("Natural Resources and Environment")
# agencytype = c("Natural Resources and Environment",
#                "Health")

# Choose Manifest Variables 
regressors = c("ma_pct", "med_sal_", "n", "appt_pct", "sup_appt_pct")
# names(fed)

# Choose Minimum Employee Size
minemployee = 5000


```

```{r}

# Data Clean
df = fed %>%
  filter(yr >= 1974) %>%
  filter(med_sal_ > 0) %>% 
  filter(n > minemployee) %>%
  drop_na(med_sal_) %>% ### Am dropping NA here, but may not need to with ctsem
  filter(AGYSUB != "TOTL") %>% ### Removes Total Rows 
  filter(!grepl("NET", agy_full)) ### Removes any total agency counts, ie only individual agencies are left

# Data Process
years = startyear:endyear
dff = df %>%
  filter(agy_typ %in% agencytype) %>%
  filter(yr %in% startyear:endyear) %>%
  select(regressors, yr, AGYSUB, agy_full) %>%
  mutate_at(regressors, scale) %>%
  drop_na() ### Again, dropping NA here, but may not need to

```

```{r}

## Make Strings for the model
## based on the number of regressors
manifest = c(0)
for (i in 2:length(regressors)) {
  temp2 = paste0("manifestmean", i)
  manifest[i] = temp2
}

```


```{r, echo = TRUE}

lambda = matrix(c(1,"lam21", "lam31", 0, 0, 0, 0 , 0, 1, "lam52"), nrow = 5, ncol = 2)


model = ctModel(type='stanct', 
               
               LAMBDA=lambda, # Forcing the first factor load to be 1
               
               n.manifest=length(regressors),
               
               manifestNames = regressors,
               
               MANIFESTMEANS = matrix(manifest,
                                      nrow = length(regressors), 
                                      ncol = 1), # Forcing the first mean to be 0
               
               n.latent=2,
               
               latentNames=c('capactiy', "autonomy"),
               
               CINT = matrix(c('cint1', "cint2"), nrow = 2),
               
               id = "AGYSUB",
               
               time = "yr")

```

```{r}

fit3 = ctStanFit(datalong = dff,
                ctstanmodel = model,
                optimize = TRUE,
                nopriors = FALSE)



```

\FloatBarrier

```{r, results='asis'}

cat(paste0('\\begin{sidewaysfigure}'))
cat(ctModelLatex(fit3,textsize = 'small',compile = FALSE, equationonly = TRUE, digits = 3))
cat('\\caption{Autonomy and Capactiy}
\\label{fig:ctmodellatexout}
\\end{sidewaysfigure}') 

```

\FloatBarrier

```{r, echo = TRUE, fig.height=5, fig.width=8}

ctKalman(fit3, 
         plot = TRUE,
         subjects = (1:length(unique(dff$AGYSUB))), # Plots all agencies
         kalmanvec = c("etasmooth", "y", "ysmooth"),
         timestep = .5) 

```

\FloatBarrier

```{r}

plot(fit3)[[1]]

```

## Interpretting Results