---
title: 'Alex Meeting 3-30-21'
author: "Blain Morin"
date: "3/30/2021"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, cache = TRUE)
library(tidyverse)
library(rstan)
library(parallel)
library(ctsem)
library(gridExtra)
library(kableExtra)
library(knitr)

```

### Overview

Here, I create some hypothetical agencies for comparison sake. 

I try the model with four variables:

* Number of employees (expected to be positively correlated with capacity)
* Median salary (expected to be positively correlated with capacity)
* Length of service (positive)
* Master's percentage (positive)

We want to see how different model specifications are effective in capturing the different agency patterns that we specify.

* **Agency A** is a large agency with increasing capacity across all measures.
* **Agency B** is a large agency with decreasing capacity across all measures.
* **Agency C** is a small agency with increasing capacity across all measures.
* **Agency D** is a medium agency that increases across measures for the first ten years, then decreases. 

### Simulated Agencies:

```{r, fig.height=7, fig.width=8}

set.seed(3322009)

yrs = 1:20

#### High capacity agency increasing
a_n = c(1000 + 50*yrs + rnorm(20, mean = 0, sd = 100))
a_sal = c(75000 + 1000*yrs + rnorm(20, mean = 0, sd = 2000))
a_los = c(10 + .1*yrs + rnorm(20, mean = 0, sd = 1))
a_ma = c(.3 + .025*yrs + rnorm(20, mean = 0, sd =.2))
a_ma = ifelse(a_ma<=0, 0, a_ma)
a_los = ifelse(a_los<=0, 0, a_los)

#### High capacity agency decreasing
b_n = c(1000 - 50*yrs + rnorm(20, mean = 0, sd = 100))
b_sal = c(75000 - 1000*yrs + rnorm(20, mean = 0, sd = 2000))
b_los = c(10 - .1*yrs + rnorm(20, mean = 0, sd = 1))
b_ma = c(.3 - .025*yrs + rnorm(20, mean = 0, sd =.2))
b_ma = ifelse(b_ma<=0, 0, b_ma)
b_los = ifelse(b_los<=0, 0, b_los)

#### Small agency increasing
c_n = c(50 + 10*yrs + rnorm(20, mean = 0, sd = 30))
c_sal = c(55000 + 500*yrs + rnorm(20, mean = 0, sd = 2000))
c_los = c(3 + .17*yrs + rnorm(20, mean = 0, sd = 1))
c_ma = c(.1 + .025*yrs + rnorm(20, mean = 0, sd =.2))
c_ma = ifelse(c_ma<=0, 0, c_ma)
c_los = ifelse(c_los<=0, 0, c_los)


#### Small agency increasing then decreasing
d_n = c(500 + 50*yrs - 3*yrs^2 + rnorm(20, mean = 0, sd = 45))
d_sal = c(55000 + 500*yrs - 15*yrs^2 + rnorm(20, mean = 0, sd = 800))
d_los = c(3 + .17*yrs - .007*yrs^2 + rnorm(20, mean = 0, sd = .3))
d_ma = c(.1 + .025*yrs - .001*yrs^2 + rnorm(20, mean = 0, sd =.01))
d_ma = ifelse(d_ma<=0, 0, d_ma)
d_los = ifelse(d_los<=0, 0, d_los)


hypdf = data.frame(AGY_sub = rep(c('a', 'b', 'c', 'd'), each = 20),
  year = rep(1:20, 4),
  n = c(a_n, b_n, c_n, d_n),
  med_sal = c(a_sal, b_sal, c_sal, d_sal),
  los = c(a_los, b_los, c_los, d_los),
  ma = c(a_ma, b_ma, c_ma, d_ma))

n.plot = hypdf %>%
  ggplot(aes(x = year, y = n)) +
  geom_point(aes(color = AGY_sub)) +
  geom_smooth(aes(color = AGY_sub)) +
  ylab("Number of Employees") +
  xlab("Year") +
  theme_minimal() +
  theme(legend.position = "top")

sal.plot = hypdf %>%
  ggplot(aes(x = year, y = med_sal)) +
  geom_point(aes(color = AGY_sub)) +
  geom_smooth(aes(color = AGY_sub)) +
  ylab("Median Salary") +
  xlab("Year") +
  theme_minimal() +
  theme(legend.position = "top")

los.plot = hypdf %>%
  ggplot(aes(x = year, y = los)) +
  geom_point(aes(color = AGY_sub)) +
  geom_smooth(aes(color = AGY_sub)) +
  ylab("Ave. Length of Service") +
  xlab("Year") +
  theme_minimal() + 
  theme(legend.position = "none")

ma.plot = hypdf %>%
  ggplot(aes(x = year, y = ma)) +
  geom_point(aes(color = AGY_sub)) +
  geom_smooth(aes(color = AGY_sub)) +
  ylab("Master's %") +
  xlab("Year") +
  theme_minimal() + 
  theme(legend.position = "none")

grid.arrange(n.plot, sal.plot,
             los.plot, ma.plot,
             nrow = 2)

```






### Two variable result (n and salary)

```{r, fig.height=7, fig.width=8, results='hide'}

set.seed(23456)

hypdf.scaled = hypdf %>%
  mutate_at(names(hypdf)[3:6], scale) %>%
  mutate(id = as.integer(as.factor(AGY_sub))) %>%
  rename(time = year)


model1 = ctModel(type='standt', LAMBDA=matrix(1, nrow = 2, ncol = 1),
               n.manifest=2, manifestNames = names(hypdf)[3:4],
               n.latent=1, latentNames=c('capacity'), CINT = "slope")

fit1 = ctStanFit(hypdf.scaled, model1, chains = 4, iterations = 2000)

#summary(fit1)

ctKalman(fit1, plot = TRUE, subjects = (1:4), kalmanvec = c("etaprior", "etasmooth", "y", "yprior")) 




```

### Four variable result

```{r, fig.height=7, fig.width=8, results='hide'}

set.seed(23)

hypdf.scaled = hypdf %>%
  mutate_at(names(hypdf)[3:6], scale) %>%
  mutate(id = as.integer(as.factor(AGY_sub))) %>%
  rename(time = year)


model2 = ctModel(type='standt', LAMBDA=matrix(1, nrow = 4, ncol = 1),
               n.manifest=4, manifestNames = names(hypdf)[3:6],
               n.latent=1, latentNames=c('capacity'), CINT = 'slope')

fit2 = ctStanFit(hypdf.scaled, model2, chains = 4, iterations = 2000)

ctKalman(fit2, plot = TRUE, subjects = (1:4), kalmanvec = c("etaprior", "etasmooth", "y", "yprior")) 

# summary(fit2)

# ctModelLatex(fit2, digits = 4, folder = "./", filename = "sim_model")

```

### Apply this model to observed agencies

```{r}

fed = read.csv("https://www.dropbox.com/s/xtt4emmz6txtbun/fed_agency_capacity_autonomy.csv?dl=1")

set.seed(123452)

### Data Clean
df = fed %>%
  filter(yr != 1973) %>%
  filter(yr >= 1974) %>%
  filter(med_sal_ > 0) %>% 
  filter(n > 5) %>%
  drop_na(med_sal_) %>%
  filter(AGYSUB != "TOTL")%>% ### Removes Total Columns 
  filter(!grepl("NET", agy_full)) %>% ### Removes any total agency counts, ie only individual agencies are left
  mutate(logMA_pct = log(ma_pct + 1)) %>%
  mutate(logLST_pct = log(LST_pct + 1)) %>%
  mutate(logAIN_pct = log(AIN_pct + 1)) %>%
  mutate(logPHDpct = log(doc_pct + 1)) %>%
  mutate(logn = log(n)) 

```

```{r, echo = TRUE}

set.seed(123452)
years = 1980:2009
setagency = c("AG16", "IN15", "EP00", "HE36")
regressors = c("n", "med_sal_", "LOSavg", "ma_pct")

```

```{r}

dff = df %>%
  filter(yr %in% years) %>%
  mutate_at(regressors, scale) %>%
  filter(AGYSUB %in% setagency) %>%
  select(regressors, yr, AGYSUB, agy_typ) %>%
  mutate(AGYSUB = as.factor(AGYSUB)) %>%
  mutate(yr = as.factor(yr)) %>%
  drop_na()

```

#### Recap 4 Agency Trends

```{r, fig.height=8, fig.width=10}

a = dff %>%
  ggplot(aes(x = yr, y = n)) +
  geom_line(aes(group = AGYSUB, color = AGYSUB)) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) + 
  ggtitle("n Employees")

b = dff %>%
  ggplot(aes(x = yr, y = med_sal_)) +
  geom_line(aes(group = AGYSUB, color = AGYSUB)) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) + 
  ggtitle("Median Salary")

c = dff %>%
  ggplot(aes(x = yr, y = LOSavg)) +
  geom_line(aes(group = AGYSUB, color = AGYSUB)) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) + 
  ggtitle("Avg Length of Service")

d = dff %>%
  ggplot(aes(x = yr, y = ma_pct)) +
  geom_line(aes(group = AGYSUB, color = AGYSUB)) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) + 
  ggtitle("Master's %")

grid.arrange(a, b, c, d, nrow = 2)

```


### Model Results

```{r, fig.height=7, fig.width=8, results='hide'}

set.seed(30)

dff = df %>%
  filter(yr %in% years) %>%
  mutate_at(regressors, scale) %>%
  filter(AGYSUB %in% setagency) %>%
  select(regressors, yr, AGYSUB, agy_typ) %>%
  mutate(AGYSUB = as.factor(AGYSUB)) %>%
  mutate(yr = as.factor(yr)) %>%
  drop_na()



model1<-ctModel(type='standt', LAMBDA=matrix(1, nrow = 4, ncol = 1),
               n.manifest=4, manifestNames=regressors,
               n.latent=1, latentNames=c('capacity'), CINT = 'slope')

dff = dff %>%
  mutate(id = as.integer(as.factor(AGYSUB))) %>%
  mutate(time = as.integer(as.factor(yr)))

fit3 = ctStanFit(dff, model1, chains = 4, iterations = 2000)

ctKalman(fit3, plot = TRUE, subjects = (1:4), kalmanvec = c("etaprior", "etasmooth", "y", "yprior")) 

```

1 = FDA, 2 = US Fish and Wildlife, 3 = Natural Resource Conservation Service, 4 = EPA